Making the data_train and data_test 


```{r}
# Set the sizes for training and testing sets
Ntrain <- 10000
Ntest <- 1000


# Generate indices for all rows
total_rows <- nrow(covertype_data)
all_indices <- 1:total_rows

# Sample indices for data_train
train_indices <- sample(all_indices, Ntrain, replace = FALSE)

# Exclude train_indices from the pool of indices
remaining_indices <- setdiff(all_indices, train_indices)

# Sample indices for data_test from the remaining indices
test_indices <- sample(remaining_indices, Ntest, replace = FALSE)

# Create data_train and data_test using the sampled indices
data_train <- covertype_data[train_indices, ]
data_test <- covertype_data[test_indices, ]

```

```{r}

# Adjust data types in data_test to match data_train
for (var in names(data_train)) {
  if (class(data_train[[var]]) != class(data_test[[var]])) {
    data_test[[var]] <- as(data_test[[var]], class(data_train[[var]]))
  }
}

# Check data types
identical_dtypes <- identical(sapply(data_train, class), sapply(data_test, class))
if (!identical_dtypes) {
  stop("Data types in data_train and data_test do not match.")
}

data_train
data_test


```

Sorting out the one hot coding in data_train 

```{r}

# Find the columns representing Wilderness_AreaX
wilderness_cols <- grep("Wilderness_Area", names(data_train))

# Get the column names without the Wilderness_Area prefix
wilderness_areas <- gsub("Wilderness_Area", "", names(data_train[wilderness_cols]))

# Find the column index where Wilderness_AreaX = 1 for each row
wilderness_area_index <- apply(data_train[, wilderness_cols], 1, function(x) which(x == 1))

# Extract the corresponding Wilderness_Area value
data_train$Wilderness_Area <- wilderness_areas[wilderness_area_index]

# Convert Wilderness_Area to a factor
data_train$Wilderness_Area <- factor(data_train$Wilderness_Area)

# Remove the original one-hot-encoded Wilderness_AreaX columns
data_train <- data_train[, -wilderness_cols]

#repeating this for the soil columns 

# Find the columns representing Soil_TypeX
soil_cols <- grep("Soil_Type", names(data_train))

# Get the column names without the Soil_Type prefix
soil_types <- gsub("Soil_Type", "", names(data_train[soil_cols]))

# Find the column index where Soil_TypeX = 1 for each row
soil_type_index <- apply(data_train[, soil_cols], 1, function(x) which(x == 1))

# Extract the corresponding Soil_Type value
data_train$Soil_Type <- soil_types[soil_type_index]

# Convert Soil_Type to a factor
data_train$Soil_Type <- factor(data_train$Soil_Type)

# Remove the original one-hot-encoded Soil_TypeX columns
data_train <- data_train[, -soil_cols]

head(data_train)


```


Running a random forest on data_train with the default parameters: 


```{r}
#Converting class to a factor so the random forest can work 
data_train$class <- factor(data_train$class)
#Running the random forest model on my data_train 
rf <- randomForest(class ~ ., data = data_train)
rf


```

Reporting variable importance: 

```{r}
imp<- importance(rf) ##extracts a matrix with the importance information
print(imp)

varImpPlot(rf) ##plots the importance information

```

Extracting the confusion matrix: 


```{r}
str(rf$confusion)


```

```{r}
#extract confusion matrix
m<- rf$confusion
#remove the last column which is the error.class.  ?
m<- m[,-ncol(m)]
m



```
```{r}
# Assuming m is your confusion matrix
m  

# Number of classes
num_classes <- nrow(m)

# Initialize vectors to store results
accuracy <- numeric(num_classes)
sensitivity <- numeric(num_classes)
specificity <- numeric(num_classes)

# Calculate metrics for each class
for (i in 1:num_classes) {
  # Accuracy
  accuracy[i] <- m[i, i] / sum(m[, i])
  
  # Sensitivity / True positive rate
  sensitivity[i] <- m[i, i] / sum(m[i, ])
  
  # True negative rate / Specificity
  true_negatives <- sum(diag(m)) - m[i, i]  # Subtract true positives from total true positives and true negatives
  total_negatives <- sum(m) - sum(m[i, ])   # Subtract true positives and false positives from total
  specificity[i] <- true_negatives / total_negatives
}

# Print or return the results
print(accuracy)
print(sensitivity)
print(specificity)

# Create a data frame with class and metrics
results_df <- data.frame(
  class = 1:num_classes,
  accuracy = accuracy,
  sensitivity = sensitivity,
  specificity = specificity
)

# Print or return the data frame
print(results_df)

```



Sorting out the one hot coding in data_test so data_train and data_test are the same 

```{r}
# Find the columns representing Wilderness_AreaX
wilderness_cols <- grep("Wilderness_Area", names(data_test))

# Get the column names without the Wilderness_Area prefix
wilderness_areas <- gsub("Wilderness_Area", "", names(data_test[wilderness_cols]))

# Find the column index where Wilderness_AreaX = 1 for each row
wilderness_area_index <- apply(data_test[, wilderness_cols], 1, function(x) which(x == 1))

# Extract the corresponding Wilderness_Area value
data_test$Wilderness_Area <- wilderness_areas[wilderness_area_index]

# Convert Wilderness_Area to a factor
data_test$Wilderness_Area <- factor(data_test$Wilderness_Area)

# Remove the original one-hot-encoded Wilderness_AreaX columns
data_test <- data_test[, -wilderness_cols]

#repeating this for the soil columns 

# Find the columns representing Soil_TypeX
soil_cols <- grep("Soil_Type", names(data_test))

# Get the column names without the Soil_Type prefix
soil_types <- gsub("Soil_Type", "", names(data_test[soil_cols]))

# Find the column index where Soil_TypeX = 1 for each row
soil_type_index <- apply(data_test[, soil_cols], 1, function(x) which(x == 1))

# Extract the corresponding Soil_Type value
data_test$Soil_Type <- soil_types[soil_type_index]

# Convert Soil_Type to a factor
data_test$Soil_Type <- factor(data_test$Soil_Type)

# Remove the original one-hot-encoded Soil_TypeX columns
data_test <- data_test[, -soil_cols]

head(data_test)


```

```{r}
# Check column names
identical_colnames <- identical(colnames(data_train), colnames(data_test))
if (!identical_colnames) {
  stop("Column names in data_train and data_test do not match.")
}

# Check and adjust factor levels if needed
for (var in names(data_test)) {
  if (is.factor(data_test[[var]])) {
    levels_train <- levels(data_train[[var]])
    levels_test <- levels(data_test[[var]])
    if (!identical(levels_train, levels_test)) {
      warning(paste("Factor levels for", var, "in data_test differ from data_train. Adjusting..."))
      data_test[[var]] <- factor(data_test[[var]], levels = levels_train)
    }
  }
}

```


Running the random forest prediction: 

```{r}
#Converting class to a factor so the random forest can work 
data_test$class <- factor(data_test$class)

#making predictions on the test data using the random forest I just made 
predictions <- predict(rf, newdata = data_test, type = "class")
#printing the predictions 
print(predictions)

```

Extracting the confusion matrix from this data

```{r}
install.packages("caret")
library(caret)
predictions_info <- confusionMatrix(table(predictions, data_test$class)) # The prediction to compute the confusion matrix and see the accuracy score
predictions_info

# Extract the confusion matrix table
m <- predictions_info$table

# Print the confusion matrix table
print(m)


#calculating the number of correctly predicted points from this 

correctly_predicted_points <- (m[1,1]) + (m[2,2]) + (m[3,3]) + (m[4,4]) + (m[5,5]) + (m[6,6]) + (m[7,7])
correctly_predicted_points 

#There are 812 correctly predicted points. 





```